Asymmetric relationships and their impact on coexistence: Supporting information

### 2023-01-25 paper/SupMat.Rnw
now done editing, for now, all parts, including the sup mat, except the Disscussion
finished editing the Discussion
minor tweaks
line numbering
edited the Intro and Fig 1 caption

\DeclareMathOperator{\EX}{E}% expected value
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\VarX}{var}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\CovX}{cov}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\mean}{mean}
\DeclareMathOperator{\se}{se}
\DeclareMathOperator{\sd}{sd}
\DeclareMathOperator{\Prob}{P}
\DeclareMathOperator{\rk}{rank}
\DeclareMathOperator{\nrk}{nrank}

@@

%header material for paper
\title{Relationships in the extremes and their influence on competition and coexistence: Supporting information}
\date{}

@@

%***DAN: Change the below paragraph if needed. Some reasons it might be needed: Steve joins as a coauthor - easy
%just means a change in wording at the top bit;
%Steve reads this and thinks it makes sense to change it because something is factually or presentationally 
%sub-optimal; or Dan or Jasmin actually develops a more rigorous understanding of scaling factors and based
%on that thinks something here is sub-optimal. Anyway, the "do both" approach seems like it can't go wrong,
%especially if Steve's prediction is correct that it won't make much difference.
%***DAN: Dan, you probably need to spend a bit more time with the cited papers to better understand this. 
We also use $q_{ij}=1$ as an alternative value, based on communication with Steve Ellner, for the following reasons.
First, as discussed in the supporting information of \cite{Ellner_2016}, and as further elaborated by 
\cite{Barabas_2018}, it takes very special circumstances for the scaling factors to exist and be unique, 
especially with size-structured models. Second, in empirically parameterized models, scaling factor values are 
often not robust to arbitrary modelling choices. 
Although the lottery model is not size structured and we do not empirically parameterize it, the above shortcomings of 
scaling factors, generally, suggest we should find alternatives to their use. The alternative 
suggested by Ellner for empirical applications is to do a series of pairwise invader-resident comparisons, each with an 
equal weighting of invader and resident $r$ values; for our context this amounts to setting scaling factors to $1$. 
We consider both approaches because the above-mentioned shortcomings do not apply to our case, but it seems 
possible that future applications will often set scaling factors to $1$, and we want our ideas to also be considered 
by researchers making that choice.

@@

%***DAN: Dan still needs to edit this section
The decomposition of storage effects and quantification of contributions of ATAs depends on 
removing asymmetric tail associations from the bivariate random variable
$(E_i,C_{i \bs i})$, while maintaining its correlation and marginal distributions, to produce a new
random variable $(E_i^\nat,C_{i \bs i}^\nat)$.
In the Theory section of the main text, we decribed this removal process
when the distributions of $E_i$ and $C_{i \bs i}$ are known exactly, but for the 
diatom example explored for goal G4 of the main text (see also SI section \ref{sect:diatom_model}), 
we only had samples from $(E_i,C_{i \bs i})$, based on simulations. We here describe how to
produce a sample from $(E_i^\nat,C_{i \bs i}^\nat)$ in this modified context. 
Let $(E_{i,k},C_{i \bs i,k})$, $k=1,\ldots,T$ denote the samples from $(E_i,C_{i \bs i})$.
Let $\rk(E_{i,k})$ be the rank of $E_{i,k}$ in the set 
of $E_{i,k}$, $k=1,\ldots,T$, where the smallest value has rank $1$. Likewise 
define $\rk(C_{i \bs i,k})$ as the rank of $C_{i \bs i,k}$ in the set
of $C_{i \bs i,k}$, $k=1,\ldots,T$. Define the ``normalized rank''
$\nrk(E_{i,k})$ as $\rk(E_{i,k})/(T+1)$, and likewise define 
$\nrk(C_{i \bs i,k})$ as $\rk(C_{i \bs i,k})/(T+1)$. Letting $\varphi$ be the CDF
of a standard normal distribution, the samples $\varphi^{-1}(\nrk(E_{i,k}))$, $k=1,\ldots,T$
are standard normally distributed, as are the samples 
$\varphi^{-1}(\nrk(C_{i \bs i,k}))$, $k=1,\ldots,T$. Let $\rho_i$ be the covariance 
of these samples. We then define a bivariate normal random variable 
$(e_i,c_{i \bs i})$ such that $\E(e_i)=\E(c_{i \bs i})=0$, $\var(e_i)=\var(c_{i \bs i})=1$,
and we denote $\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We used $\ddot{\rho}_i = \rho_i$.
Generating a sample $(e_{i,k},c_{i \bs i,k})$, $k=1,\ldots,T$ from this distribution,
we permute the ordered set $(E_{i,1},E_{i,2},\ldots,E_{i,T})$ so that the ranks of the 
permuted ordered set equal the ranks of the corresponding elements of the ordered
set $(e_{i,1},e_{i,2},\ldots,e_{i,T})$. We likewise permute the ordered set 
$(C_{i \bs i,1},C_{i \bs i,2},\ldots,C_{i \bs i,T})$ so the ranks of the permuted result
match the ranks of $(c_{i \bs i,1},c_{i \bs i,2},\ldots,c_{i \bs i,T})$.
This step works as long as there are no ties among the $E_{i,k}$ or among
the $C_{i \bs i,k}$. There were no ties when we applied these methods to the diatom system. 
The two permuted ordered sets, here denoted $(E_{i,1}^\nat,E_{i,2}^\nat,\ldots,E_{i,T}^\nat)$
and $(C_{i \bs i,1}^\nat,C_{i \bs i,2}^\nat,\ldots,C_{i \bs i,T}^\nat)$,
when combined to make ordered pairs, form the desired sample from 
$(E_i^\nat,C_{i \bs i}^\nat)$. Because $(E_{i,1}^\nat,\ldots,E_{i,T}^\nat)$
is a permutation of $(E_{i,1},\ldots,E_{i,T})$ and 
$(C_{i \bs i,1}^\nat,\ldots,C_{i \bs i,T}^\nat)$ is a permutation of
$(C_{i \bs i,1},\ldots,C_{i \bs i,T})$, marginal distributions of our
sample from $(E_i^\nat,C_{i \bs i}^\nat)$ were the same as those of our original 
sample from $(E_i, C_{i \bs i})$. The choice of bivariate normal
$(e_i,c_{i \bs i})$ ensures that our resulting sample shows symmetric tail association.
Finally, we caution that sampling variation means any quantity computed from our resulting sample 
$(E_{i,k}^\nat,C_{i \bs i,k}^\nat)$, $k=1,\ldots,T$ should be recomputed for 
multiple such samples, all computed via the same process, and a median of 
the resulting distribution of values should be used.

>>>>

\DeclareMathOperator{\E}{\mathbb{E}}% expected value
\DeclareMathOperator{\VarX}{var}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\CovX}{cov}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\mean}{mean}
\DeclareMathOperator{\se}{se}
\DeclareMathOperator{\sd}{sd}
\DeclareMathOperator{\Prob}{P}
\DeclareMathOperator{\rk}{rank}
\DeclareMathOperator{\nrk}{nrank}

@@

%header material for paper
\title{Asymmetric relationships and their impact on coexistence: Supporting information}
\date{}

@@

%***DAN: Change the below paragraph if needed. Some reasons it might be needed: Steve joins as a coauthor - easy
%just means a change in wording at the top bit;
%Steve reads this and thinks it makes sense to change it because something is factually or presentationally 
%sub-optimal; or Dan or Jasmin actually develops a more rigorous understanding of scaling factors and based
%on that thinks something here is sub-optimal. Anyway, the "do both" approach seems like it can't go wrong,
%especially if Steve's prediction is correct that it won't make much difference.
%***DAN: Dan, you probably need to spend a bit more time with the cited papers to better understand this. 
We also use $q_{ij}=1$ as an alternative value, for the following reasons.
First, as discussed in the supporting information of \cite{Ellner_2016}, and as further elaborated by 
\cite{Barabas_2018}, it takes very special circumstances for the scaling factors to exist and be unique, 
especially with size-structured models. Second, in empirically parameterized models, scaling factor values are 
often not robust to arbitrary modelling choices. 
Although the lottery model is not size structured and we do not empirically parameterize it, the above shortcomings of 
scaling factors, generally, suggest we should find alternatives to their use. The alternative 
suggested by Ellner for empirical applications is to do a series of pairwise invader-resident comparisons, each with an 
equal weighting of invader and resident $r$ values; for our context this amounts to setting scaling factors to $1$. 
We consider both approaches because the above-mentioned shortcomings do not apply to our case, but it seems 
possible that future applications will often set scaling factors to $1$, and we want our ideas to also be considered 
by researchers making that choice.

@@

The decomposition of storage effects and quantification of contributions of ATAs depends on 
removing asymmetric tail associations from the bivariate random variable
$(E_i,C_{i \bs i})$, while maintaining its correlation and marginal distributions, to produce a new
random variable $(E_i^\nat,C_{i \bs i}^\nat)$.
In the Theory section of the main text, we decribed this removal process
when the distributions of $E_i$ and $C_{i \bs i}$ are known exactly, but for the 
diatom example explored for goal G4 of the main text (see also SI section \ref{sect:diatom_model}), 
we only had samples from $(E_i,C_{i \bs i})$, based on simulations. We here describe how to
produce a sample from $(E_i^\nat,C_{i \bs i}^\nat)$ in this modified context. 
Let $(E_{i,k},C_{i \bs i,k})$, $k=1,\ldots,T$ denote the samples from $(E_i,C_{i \bs i})$.
Let $\rk(E_{i,k})$ be the rank of $E_{i,k}$ in the set 
of $E_{i,k}$, $k=1,\ldots,T$, where the smallest value has rank $1$. Likewise 
define $\rk(C_{i \bs i,k})$ as the rank of $C_{i \bs i,k}$ in the set
of $C_{i \bs i,k}$, $k=1,\ldots,T$. Define the ``normalized rank''
$\nrk(E_{i,k})$ as $\rk(E_{i,k})/(T+1)$, and likewise define 
$\nrk(C_{i \bs i,k})$ as $\rk(C_{i \bs i,k})/(T+1)$. Letting $\varphi$ be the CDF
of a standard normal distribution, the samples $\varphi^{-1}(\nrk(E_{i,k}))$, $k=1,\ldots,T$
are standard normally distributed, as are the samples 
$\varphi^{-1}(\nrk(C_{i \bs i,k}))$, $k=1,\ldots,T$. Let $\rho_i$ be the covariance 
of these samples. We then define a bivariate normal random variable 
$(e_i,c_{i \bs i})$ such that $\E(e_i)=\E(c_{i \bs i})=0$, $\var(e_i)=\var(c_{i \bs i})=1$,
and we denote $\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We used $\ddot{\rho}_i = \rho_i$.
Generating a sample $(e_{i,k},c_{i \bs i,k})$, $k=1,\ldots,T$ from this distribution,
we permute the ordered set $(E_{i,1},E_{i,2},\ldots,E_{i,T})$ so that the ranks of the 
permuted ordered set equal the ranks of the corresponding elements of the ordered
set $(e_{i,1},e_{i,2},\ldots,e_{i,T})$. We likewise permute the ordered set 
$(C_{i \bs i,1},C_{i \bs i,2},\ldots,C_{i \bs i,T})$ so the ranks of the permuted result
match the ranks of $(c_{i \bs i,1},c_{i \bs i,2},\ldots,c_{i \bs i,T})$.
This step works as long as there are no ties among the $E_{i,k}$ or among
the $C_{i \bs i,k}$. There were no ties when we applied these methods to the diatom system. 
The two permuted ordered sets, here denoted $(E_{i,1}^\nat,E_{i,2}^\nat,\ldots,E_{i,T}^\nat)$
and $(C_{i \bs i,1}^\nat,C_{i \bs i,2}^\nat,\ldots,C_{i \bs i,T}^\nat)$,
when combined to make ordered pairs, form the desired sample from 
$(E_i^\nat,C_{i \bs i}^\nat)$. Because $(E_{i,1}^\nat,\ldots,E_{i,T}^\nat)$
is a permutation of $(E_{i,1},\ldots,E_{i,T})$ and 
$(C_{i \bs i,1}^\nat,\ldots,C_{i \bs i,T}^\nat)$ is a permutation of
$(C_{i \bs i,1},\ldots,C_{i \bs i,T})$, marginal distributions of our
sample from $(E_i^\nat,C_{i \bs i}^\nat)$ were the same as those of our original 
sample from $(E_i, C_{i \bs i})$. The choice of bivariate normal
$(e_i,c_{i \bs i})$ ensures that our resulting sample shows symmetric tail association.
Finally, we caution that sampling variation means any quantity computed from our resulting sample 
$(E_{i,k}^\nat,C_{i \bs i,k}^\nat)$, $k=1,\ldots,T$ should be recomputed for 
multiple such samples, all computed via the same process, and a median of 
the resulting distribution of values should be used.
